Disk Winds Matter: Simulating the Optical Spectra of Cataclysmic Variables

### 2014-08-22 draft.tex
Added figure captions and a bit more text

%% define some control sequences for lines
\def\heiiopt{He~\textsc{ii}~$\lambda4686{\rm \AA}$}
\def\heiiuv{He~\textsc{ii}~$\lambda1640{\rm \AA}$}
\def\heiioptnew{He~\textsc{ii}~$\lambda3202{\rm \AA}$}
\def\la{Ly-$\alpha$}
\def\ha{H$\alpha$}
\def\hb{H$\beta$}

@@

\py has previously been tested against a number of radiative transfer and
photoionization codes. Comparisons of ionization balance with \cld \citep{cloudy2013} 
can be found in LK02 and H13 showing excellent agreement. We have also conducted comparisons
of ionization and spectral synthesis with the supernova code \textsc{Tardis.} \tar is described by
\cite{kerzendorfsim}, and the spectral comparisons can be found therein.
In addition to these code validation efforts, we have also conducted tests of
the macro atom scheme in \py to verify that it does indeed solve for level populations correctly
and produces the correct level emissivities. We have been able to show good agreement with 
Case A and Case B calculations by \cite{seaton1959}, and have tested our level
population solver against \textsc{Tardis.}

@@

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/acc_law.eps}
\caption{
The velocity law in our model for various values of
the acceleration exponent.
}
\label{acc_law}
\end{figure}

@@

\begin{figure} %fullpage
\includegraphics[width=\textwidth]{figures/fig5_uv_opt.eps}
\caption{
UV (left) and Optical synthetic spectra for model B computed at
sightlines of 10,27.5, 45, 62.5 and 80 degrees.	
The inset plots show zoomed-in line profiles for 
\heiiuv and \ha. Double-peaked line emission can be seen in 
\heiiuv, \heiiopt, \ha and some He I lines, but the 
line emission is not always sufficient to overcome the absorption
cores from the stellar atmosphere models.
}
\label{spec}
\end{figure} %fullpage\

@@

\begin{figure} 
\includegraphics[width=0.45\textwidth]{figures/fig6_opt_cont.eps}
\caption{Synthetic optical spectra computed for 
sightlines of 22.5, 25, 62.5 and 80 degrees.}
\label{spec_continuum}
\end{figure}

@@

\begin{figure} %fullpage
\includegraphics[width=\textwidth]{figures/fig5.eps}
\caption{Various properties of the wind.}
\label{wind}
\end{figure} %fullpage

@@

We introduce blackbody boundary layers with temperatures of 80,000K and 200,000K,
commonly proposed values in the literature, and adopt a scale height according to
equation~\ref{bl}. We find that the strength of the C\textsc{iv} line decreases
with increasing boundary layer temperature, as shown in figure~\ref{jump}.
We note that a model boundary layer could produce a softer
spectrum for a given effective temperature \citep[e.g.][]{suleimanov2014}, 
meaning that our ionization constraints would
allow for a higher value of $T_{bl}$. It is also
possible that a flatter temperature profile in the disk,
as suggested by some eclipse mapping observations of NLs (References),
would cause a softer disk spectrum and allow a boundary layer 
with $T_{bl} = 200,000$K to exist without over-ionizing the wind.

@@

@@

\begin{figure} %fullpage
\includegraphics[width=\textwidth]{figures/fig14_uv_opt.eps}
\caption{
UV (left) and Optical synthetic spectra for model B computed at
sightlines of 10,27.5, 45, 62.5 and 80 degrees.	
The inset plots show zoomed-in line profiles for 
\heiiuv \ and \ha. The \ha line 
is single peaked, but higher order lines in the Balmer series
are double-peaked, albeit with narrower profiles.
Strong \heiiopt emission can be seen
}
\label{uvoptb}
\end{figure} %fullpage

@@

\renewcommand{\labelitemi}{$\bullet$}
\begin{itemize}
	\item The `filling-in' of the Balmer jump \citep{elitzur1983} by recombination 
	continuum emission from a disk wind
	\item Single peaked line emission in H alpha produced by 
	the introduction of a poloidal velocity gradient \citep{MC96}
	\item A Balmer decrement consistent with optically thick emission \citep{elitzur1983}.
\end{itemize}
\smallskip

>>>>

%% define some control sequences for lines
\def\heiiopt{He~\textsc{ii}~$\lambda4686{\rm \AA}$}
\def\heiiuv{He~\textsc{ii}~$\lambda1640{\rm \AA}$}
\def\heiioptnew{He~\textsc{ii}~$\lambda3202{\rm \AA}$}
\def\la{Ly-$\alpha$}
\def\ha{H$\alpha$}
\def\hb{H$\beta$}
\def\civfull{C~\textsc{iv}~$\lambda1550{\rm \AA}$}

@@

\py has previously been tested against a number of radiative transfer and
photoionization codes. Comparisons of ionization balance with \cld \citep{cloudy2013} 
can be found in LK02 and H13 showing excellent agreement. We have also conducted comparisons
of ionization and spectral synthesis with the supernova code \textsc{Tardis.} \tar is described by
\cite{kerzendorfsim}, and the spectral comparisons can be found therein.
In addition to these code validation efforts, we have also conducted tests of
the macro atom scheme in \py to verify that it does indeed solve for level populations correctly
and produces the expected level emissivities. Figure~\ref{tests} shows good agreement with 
Case B, l-mixed calculations by \cite{seaton1959}, and level
population calcaultions with \textsc{Tardis.}

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/fig_caseb_tardis.eps}
\caption{
{\sl Top Panel:} Balmer decrements from \py compared with analytic calculations
by Seaton (1959), showing good agreement when calculating level emissivities in Hydrogen
in the Case B limit.
(see Osterbrock 1989 for a discussion of this commonly used approximation).
{\sl Bottom Panel:}  a comparison of Helium I level populations (the most complex ion we currently 
treat as a macro-atom) between \py~and \tar~models. Considering the 
two codes use different atomic data and \textsc{Tardis,} unlike \textsc{Python,} currently has 
a complete treatment of collisions between radiatively forbidden transitions, the factor of 
$<2$ agreement is encouraging. 
}
\label{tests}
\end{figure}
\nocite{osterbrock}
\nocite{seaton1959}

@@

\begin{figure}
\centering
\includegraphics[width=0.45\textwidth]{figures/acc_law.eps}
\caption{
The velocity law in our model for various values of
the acceleration exponent, $\alpha$.
}
\label{acc_law}
\end{figure}

@@

\begin{figure} %fullpage
\includegraphics[width=\textwidth]{figures/fig5_uv_opt.eps}
\caption{
UV (left) and Optical synthetic spectra for model A, our benchmark model,
computed at sightlines of 10, 27.5, 45, 62.5 and 80 degrees.	
The inset plots show zoomed-in line profiles for 
\heiiuv and \ha. Double-peaked line emission can be seen in 
\heiiuv, \heiiopt, \ha\ and some He I lines, but the 
line emission is not always sufficient to overcome the absorption
cores from the stellar atmosphere models. The model
also produces a prominent \heiioptnew\ line at high inclinations,
}
\label{spec}
\end{figure} %fullpage\

@@

\begin{figure} 
\includegraphics[width=0.45\textwidth]{figures/fig6_opt_cont.eps}
\caption{Synthetic optical spectra computed for 
sightlines of 22.5, 25, 62.5 and 80 degrees. In these plots
the flux is divided by a polynomial fit to the 
underlying continuum redward of the Balmer edge, so that 
line-to-continuum ratios and the true depth of the
Balmer jump can be shown.}
\label{spec_continuum}
\end{figure}

@@

\begin{figure} %fullpage
\includegraphics[width=\textwidth]{figures/fig5.eps}
\caption{
The Physical Properties of the wind. Near
the disk plane the wind is dense, with low poloidal velocities.
As the wind accelerates it becomes less dense
and more highly ionized. The dominant Helium ion
is almost always He III, apart from in a small
portion of the wind at the base, which is partially shielded
from the inner disk.
}
\label{wind}
\end{figure} %fullpage

@@

We introduce blackbody boundary layers with temperatures of 80,000K and 200,000K,
commonly proposed values in the literature, and adopt a scale height according to
equation~\ref{bl}. We find that the strength of the C\textsc{iv} line decreases
with increasing boundary layer temperature, as shown in figure~\ref{jump}.
We note that a model boundary layer could produce a softer
spectrum for a given effective temperature \citep[e.g.][]{suleimanov2014}, 
meaning that our ionization constraints would
allow for a higher value of $T_{bl}$. It is also
possible that a flatter temperature profile in the disk,
as suggested by some eclipse mapping observations of NLs (References),
would cause a softer disk spectrum and allow a boundary layer 
with $T_{bl} = 200,000$K to exist without over-ionizing the wind.
We note that there is an element of inconsistency with 
requiring flatter disk temperature profiles in order to permit
a hotter boundary layer,, as the theoretical argument 
for the existence and likely temperature of a boundary layer has 
mostly been put together assuming a Shakura-Sunyaev disk.  
Radio emission has been seen in CVs, possibly implying the 
existence of a radio jet \cite[see e.g.][]{kordingDNjet2008,kording2011}.
A jet could extract some of the energy which is supposedly dissipated in the 
boundary layer, causing $L_{bl} < L_{acc}/2$, although
observations suggest the jet power would only be on the order
of $10\%$ of the accretion power.

@@

\begin{figure} 
\mbox{
\subfigure{\includegraphics[width=0.5\textwidth]{figures/fig_EWt.eps}}

}
%%\includegraphics[width=0.5\textwidth]{figures/test.eps}
\caption{
Equivalent widths of the \civfull\ and \heiiuv\ lines
as a function of boundary layer temperature. The models used
are identical to the benchmark model in geometry and kinematics.
% The panels are ordered in the same way as figure~\ref{halpha}. In general, increasing densities
% lead to a Balmer jump in emission, although as in the case of H$\alpha$ there is a transition
% when the wind becomes a neutral absorbing column.
}
\label{bl_grid}
\end{figure}

@@

\begin{figure} %fullpage
\includegraphics[width=\textwidth]{figures/fig14_uv_opt.eps}
\caption{
UV (left) and Optical synthetic spectra for model B computed at
sightlines of 10, 27.5, 45, 62.5 and 80 degrees.	
The inset plots show zoomed-in line profiles for 
\heiiuv \ and \ha. The \ha line 
is single peaked, but higher order lines in the Balmer series
are double-peaked, albeit with narrower profiles.
Strong \heiiopt emission can be seen, as well a trend
of a deeper Balemr jump with decreasing inclination.
}
\label{uvoptb}
\end{figure} %fullpage

@@

\renewcommand{\labelitemi}{$\bullet$}
\begin{itemize}
	\item The `filling-in' of the Balmer jump \citep{elitzur1983} by recombination 
	continuum emission from a disk wind
	\item Single peaked line emission in \ha\ produced by 
	the introduction of a poloidal velocity gradient \citep{MC96}
	\item A Balmer decrement consistent with optically thick emission \citep{elitzur1983}.
\end{itemize}
\smallskip
